Reasonably Programmable Syntax

### 2017-02-28 syntax-icfp17.tex
ICFP 2017 submission

%% For double-blind review submission
\documentclass[acmlarge,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission
%\documentclass[acmlarge,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission
%\documentclass[acmlarge]{acmart}\settopmatter{}

@@

%% Cyrus packages
\usepackage{microtype}

%% Listings
\usepackage{listings}
\lstset{tabsize=2, 
basicstyle=\ttfamily\fontsize{7.5pt}{1em}\selectfont, 
commentstyle=\itshape\ttfamily\color{gray}, 
stringstyle=\ttfamily\color{red},
mathescape=false,escapechar=!,
numbers=left, numberstyle=\scriptsize\color{mygray}, language=ML,moredelim=[il][\sffamily]{?},showspaces=false,showstringspaces=false,xleftmargin=0pt, numbersep=3pt, morekeywords=[1]{tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,datatype,new,toast,syntax,module,where,import,for,ana,syn,opcon,tycon,metasignature,metamodule,metasig,metamod,static,at,by,tycase,mod,macro,match,pattern,in,patterns,expressions,implicit,forall,exptsm,pattsm},deletekeywords={double},classoffset=0,
aboveskip=2pt,belowskip=0pt,
moredelim=**[is][\color{red}]{SSTR}{ESTR},
moredelim=**[is][\color{OliveGreen}]{SHTML}{EHTML},
moredelim=**[is][\color{purple}]{SCSS}{ECSS},
moredelim=**[is][\color{brown}]{SSQL}{ESQL},
moredelim=**[is][\color{orange}]{SCOLOR}{ECOLOR},
moredelim=**[is][\color{magenta}]{SPCT}{EPCT}, 
moredelim=**[is][\color{gray}]{SNAT}{ENAT}, 
moredelim=**[is][\color{blue}]{SURL}{EURL},
moredelim=**[is][\color{SeaGreen}]{SQT}{EQT},
moredelim=**[is][\color{Periwinkle}]{SGRM}{EGRM},
moredelim=**[is][\color{YellowGreen}]{SID}{EID},
moredelim=**[is][\color{Sepia}]{SUS}{EUS},
deletestring=[d]{"},
}
\newcommand{\liv}[1]{\lstinline{#1}}
\newcommand{\li}[1]{\lstinline{#1}}

@@

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
Programmers, 
like ``pen-and-paper'' mathematicians, often seek to define ``syntactic sugar'' that  lowers the syntactic cost of common idioms. 
Unfortunately, the most direct mechanisms available to programmers, e.g. \li{camlp4} \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, do not support modular reasoning about determinism (i.e. separately defined forms can conflict syntactically with one another), and they obscure the type and binding structure of the program. We argue therefore that these mechanisms are unreasonable for programming ``in the large''.

This paper formally introduces \emph{parametric typed syntax macros} (pTSMs), which give library providers programmatic control over both the parsing and expansion of expressions and patterns of \emph{generalized literal form} at a specified type or parameterized family of types. Expansion is \emph{strictly hygienic}, meaning that expansions can refer to external bindings only via spliced terms or explicit module parameters, and expansions do not reveal internal bindings to spliced terms or to the remainder of the program. 
% Partial parameter application lowers the syntactic cost of this strict style. 
This design ensures that clients are able to reason about the binding structure of the program while holding the expansion of the program abstract. The system needs only convey to clients a \emph{segmentation} of each literal that gives the locations of the spliced terms. The system can determine this {segmentation} automatically and convey it to the client via syntax highlighting, or by presenting the client with a context-free grammar (without revealing the semantic actions associated with each production.) 
We argue that this mechanism occupies a ``sweet spot'' in the design space, in that it captures many common syntactic idioms while avoiding the problem of syntactic conflicts by construction and supplying clients with clear abstract reasoning principles.

@@

\section{Introduction}

Text of paper \ldots

>>>>

\PassOptionsToPackage{svgnames,dvipsnames,svgnames}{xcolor}
%% For double-blind review submission
\documentclass[acmlarge,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission
%\documentclass[acmlarge,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission
%\documentclass[acmlarge]{acmart}\settopmatter{}

@@

%% Cyrus packages
\usepackage{microtype}
\usepackage{todonotes}
\usepackage{mdframed}
\usepackage{colortab}
\usepackage{mathpartir}
\usepackage{enumitem}
\usepackage{bbm}

% \newtheorem{theorem}{Theorem}[chapter]
% \newtheorem{lemma}[theorem]{Lemma}
% \newtheorem{corollary}[theorem]{Corollary}
% \newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem{condition}[theorem]{Condition}

%% Listings
\definecolor{mygray}{rgb}{0.75,0.75,0.75}
\usepackage{listings}
\lstset{tabsize=2, 
basicstyle=\ttfamily\fontsize{10pt}{1em}\selectfont, 
commentstyle=\itshape\ttfamily\color{gray}, 
stringstyle=\ttfamily\color{red},
mathescape=false,escapechar=~,
numbers=left, numberstyle=\scriptsize\color{mygray}, language=ML,moredelim=[il][\sffamily]{?},showspaces=false,showstringspaces=false,xleftmargin=0pt, numbersep=3pt, morekeywords=[1]{spliced,tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,datatype,new,toast,syntax,module,where,import,for,ana,syn,opcon,tycon,metasignature,metamodule,metasig,metamod,static,at,by,tycase,mod,macro,match,pattern,in,patterns,expressions,implicit,forall,exptsm,pattsm},deletekeywords={double},classoffset=0,
aboveskip=4pt,belowskip=2pt,
moredelim=**[is][\color{red}]{SSTR}{ESTR},
moredelim=**[is][\color{OliveGreen}]{SHTML}{EHTML},
moredelim=**[is][\color{purple}]{SCSS}{ECSS},
moredelim=**[is][\color{brown}]{SSQL}{ESQL},
moredelim=**[is][\color{orange}]{SCOLOR}{ECOLOR},
moredelim=**[is][\color{magenta}]{SPCT}{EPCT}, 
moredelim=**[is][\color{gray}]{SNAT}{ENAT}, 
moredelim=**[is][\color{OliveGreen}]{SURL}{EURL},
moredelim=**[is][\color{SeaGreen}]{SQT}{EQT},
moredelim=**[is][\color{Periwinkle}]{SGRM}{EGRM},
moredelim=**[is][\color{YellowGreen}]{SID}{EID},
moredelim=**[is][\color{Sepia}]{SUS}{EUS},
deletestring=[d]{"},
}
\newcommand{\liv}[1]{\lstinline{#1}}
\newcommand{\li}[1]{\lstinline{#1}}

@@

\input{../../thesis/macros}

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
Programmers, 
like pen-and-paper mathematicians, often wish to define  ``syntactic sugar'' to  lower the syntactic cost of common idioms. 
Unfortunately, the most direct mechanisms of syntactic control available to programmers, e.g. \li{camlp4} \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, do not support modular reasoning about determinism (i.e. separately defined forms can conflict syntactically with one another.) Moreover, they obscure the type and binding structure of the program. Typed hygienic term-rewriting macros (e.g. as available in Scala) are more reasonable, but they give programmers only limited syntactic control.

This paper formally introduces \emph{simple} and \emph{parametric typed syntax macros} (TSMs), which give library providers programmatic control over both the parsing and expansion of expressions and patterns of \emph{generalized literal form} at a specified type or parameterized family of types. Expansion is {strictly hygienic}, meaning that 1) expansions can refer to external bindings only via spliced terms or supplied module parameters; and 2)  expansions do not reveal internal bindings to spliced terms or to the remainder of the program. 
% Partial parameter application lowers the syntactic cost of this strict style. 
This design ensures that clients are able to reason about the binding structure of the program while holding the expansion of the program abstract -- the system needs only convey to clients the \emph{segmentation} of each literal, which gives the locations of the spliced terms. 
%The system can determine this {segmentation} automatically and convey it to the client via syntax highlighting, or by presenting the client with a context-free grammar (without revealing the semantic actions associated with each production.) 
We argue that this mechanism occupies a ``sweet spot'' in the design space, in that it captures many common syntactic idioms while avoiding the problem of syntactic conflicts by construction and supplying clients with clear abstract reasoning principles. We establish these reasoning principles formally.

@@

\newcommand{\introSec}{Introduction}
\section{\protect\introSec}
\label{sec:intro}
When defining the textual syntax of a programming language, it is common practice to include ``syntactic sugar'' specific to certain standard library constructs.   
For example, Standard ML and other functional languages include sugar for constructing and pattern matching on lists, as defined in the standard library \cite{mthm97-for-dart}.

Third-party library providers are, quite justifiably, envious of the privileged status of these otherwise seemingly ordinary standard library definitions. After all, it is not difficult to find other examples of situationally useful library-specific syntactic sugar \cite{TSLs}. For example, 1) clients of a ``collections'' library might like syntactic sugar for finite sets and dictionaries; 2) clients of a ``web programming'' library might like syntactic sugar for HTML and JSON values; 3) a compiler writer might like syntactic sugar for the terms of the object language or various intermediate languages of interest; and 4) clients of a ``chemistry'' library might like syntactic sugar for chemical structures based on the SMILES standard \cite{anderson1987smiles}. This has motivated a long line of research into approaches that delegate more syntactic control to library providers.

\subsection{Syntax Dialects are Unreasonable}\label{sec:syntax-dialects-intro}
One well-worn approach is for library providers to use a syntax definition system to construct a library-specific \emph{syntax dialect}, i.e. a new syntax definition that extends the syntax definition given in the language definition with new derived forms (i.e. forms given meaning by expansion, rather than directly.) For example, Ur/Web defines a dialect of Ur's textual syntax with derived forms for SQL queries, XHTML elements and some other constructs defined in a  web programming library \cite{conf/popl/Chlipala15,conf/pldi/Chlipala10}. Figure \ref{fig:urweb} demonstrates how an XHTML expression containing ``spliced'' string expressions (Line 1) and XHTML expressions (Line 2) can be expressed in Ur/Web. 
The desugarings of these derived forms, not shown, are substantially more verbose and, for programmers familiar with the standardized syntax for XHTML \cite{xhtml}, substantially more obscure.

\begin{figure}
\begin{lstlisting}
  fun heading first_name = SURL<xml><h1>Hello, {[EURLfirst_nameSURL]}!</h1></xml>EURL
  val body = SURL<xml><body>{EURLheading "World"SURL} ...</body></xml>EURL
\end{lstlisting}
\vspace{-6px}
\caption{Derived XHTML forms with support for splicing at two different types are built in to Ur/Web.}
\vspace{-6px}
\label{fig:urweb}
\end{figure}
 % Such dialects are sometimes qualitatively taxonomized as amongst the ``domain-specific language'' for this reason \cite{fowler2010domain}. %Syntactic cost is often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined. 
% Syntax definition systems , which we will discuss in Sec. \ref{sec:syntax-dialects}, have simplified the task of defining ``library-specific'' (a.k.a. ``domain-specific'') syntax dialects like Ur/Web, and have thereby contributed to their ongoing proliferation.

% Many have argued that a proliferation of syntax dialects constructed using these tools is harmless or even desirable, because programmers can simply choose the right syntax dialect for each job at hand \cite{journals/stp/Ward94}. However, we argue that in fact 
There are hundreds of syntax definition systems of various design, e.g. grammar-based systems like Camlp4 \cite{ocaml-manual}, Copper \cite{conf/gpce/WykS07} or SugarJ/Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, parser combinator systems \cite{Hutton1992d} and somewhat less general systems built in to certain programming systems, like the system of fixity directives in Standard ML \cite{mthm97-for-dart} or more elaborate mixfix systems \cite{5134,DBLP:conf/gpce/TahaJ03} built in to proof assistants like Agda \cite{DBLP:conf/ifl/DanielssonN08} and Coq \cite{Coq:manual}. 
%library-specific (a.k.a. ``domain-specific'') 
Unfortunately, this dialect-oriented approach is difficult to reconcile with the best practices of ``programming in the large'', i.e. developing large programs ``consisting of many small programs (modules), possibly written by different people''   \cite{DeRemer76} whose interactions are mediated by a reasonable type and binding discipline. The problems that tend to arise are summarized below.%a more systematic treatment will follow in  Sec. \ref{sec:syntax-dialects}.

\paragraph{Problem 1: Incompatibility and Syntactic Conflict}
The first problem with the dialect-oriented approach is that clients  cannot always combine different syntax dialects when they want to use derived forms that they define together. This is a significant problem because large programs use many separately defined libraries \cite{DBLP:conf/sac/LammelPS11} and often do not fall cleanly into a single preconceived ``problem domain''.

In some cases, the dialects in question cannot be combined simply because they have been defined under incompatible syntax definition systems. In other cases, there is no dialect combination operation available that ensures that the combined dialect is an extension of both constituent dialects. And in other cases, the problem is one of \emph{conflict}: two dialects might define the same form, but determine different desugarings.

For example, consider two syntax dialects defined under a system like Camlp4: $\mathcal{D}_1$ defines derived forms for sets, and $\mathcal{D}_2$ defines derived forms for dictionaries, both delimited by \verb~{<~ and \verb~>}~.\footnote{In OCaml, simple curly braces are already reserved by the language for record types and values.} Each dialect is deterministic (i.e. there are no possible syntactic conflicts), $\mathrm{det}(\mathcal{D}_1)$ and $\mathrm{det}(\mathcal{D}_2)$, but when the grammars are combined by Camlp4, it is not the case that $\mathrm{det}(\mathcal{D}_1 \cup \mathcal{D}_2)$. In particular, \verb~{<>}~ can desugar to the empty set and the empty dictionary.%(Indeed, this ambiguity arose when Python added set literals to its grammar.) %A recent version of Python added derived forms for mutable sets. Due to a conflict with dictionary syntax, however, there is no derived form for the empty set.)
 %A third syntax dialect might come along that uses the same forms that $\mathcal{D}_2$ defines, but for ordered finite maps.

\citet{conf/pldi/SchwerdfegerW09} have developed a modular grammar analysis, implemented in Copper, that ``nearly'' guarantees that determinism is conserved when syntax dialects (of a certain restricted class) are combined, the caveat being that the constituent dialects must prefix all newly introduced forms with marking tokens drawn from disjoint sets. To be confident that the marking tokens used are disjoint, providers must base them on the domain name system (or some other coordinating entity.) Because the mechanism operates at the level of the context-free grammar, it is difficult for the client to define scoped abbreviations for these verbose marking tokens (as one might do when, for example, deeply nested modules are used for namespacing in ML.)

\paragraph{Problem 2: Abstract Reasoning About Derived Forms}

Even putting aside the problem of syntactic conflict, there are questions about how \emph{reasonable}  sprinkling library-specific derived forms throughout a large software system might be. 
For example, consider the perspective of a programmer attempting to comprehend (i.e. reason about) the program fragment in Figure \ref{fig:K-dialect}, which is written in a syntax dialect constructed by combining various library-specific dialects of Standard ML's syntax.

\begin{figure}
\begin{lstlisting}
  val w = compute_w ()
  val x = compute_x ()
  val y = {|(!R)@&{&/x!/:2_!x}'!R}|}
\end{lstlisting}
\vspace{-8px}
\caption{An example of unreasonable program text.}
\vspace{-8px}
\label{fig:K-dialect}
\end{figure}

If the programmer happens to be familiar with the (intentionally terse) syntax of the stack-based database query processing language K \cite{Whitney:2001:LOR:376284.375783}, then Line 3 might pose few difficulties. If the programmer does not recognize this syntax, however, there are no simple, definitive protocols for answering questions like:

\begin{enumerate}
\item \textbf{(Responsibility)} Which constituent dialect defined the derived form that appears on Line 3?
\item \textbf{(Segmentation)} Are the characters \li{x} and \li{R} on Line 3 parsed as spliced expressions \li{x} and \li{R} (i.e. expressions of variable form), or parsed in some other way peculiar to this form?
\item \textbf{(Capture)} If \li{x} is in fact a spliced expression, does it refer to the binding of \li{x} on Line 2? Or might it capture an unseen binding introduced in the desugaring of Line 3?
\item \textbf{(Context Dependence)} If \li{w}, on Line 1, is renamed, could that possibly break the program, or change its meaning? In other words, might the desugaring of Line 3 assume that some variable identified as \li{w} is in scope (even though \li{w} is not mentioned in the text of Line 3)?
\item \textbf{(Typing)} What type does \li{y} have?
\end{enumerate}

In short, syntax dialects do not come with reasonable principles of \emph{syntactic abstraction}: if the desugaring of the program text is held abstract, programmers can no longer reason about types and binding (i.e. answer questions like those above) in the usual disciplined manner. This is burdensome at all scales, but particularly when programming in the large, where it is common to encounter unfamiliar libraries. Forcing the programmer to examine the desugaring of the program text in order to reason about types and binding defeats the ultimate purpose of syntactic sugar -- decreasing cognitive cost \cite{Green89}.

%In other words, encountering an unfamiliar derived form has made it difficult for the programmer to maintain the usual \emph{type discipline} and \emph{binding discipline}. %Compelling the programmer to examine the desugaring directly defeat the purpose of defining the derived form -- decreasing cognitive cost. Indeed, it substantially increases cognitive cost.

% In contrast, when a programmer encounters, for example, a function call like the call to \li{read_data} on Line 3, the analagous questions can be answered by following clear protocols that become ``cognitive reflexes'' after sufficient experience with the language, even if the programmer has no experience with the library defining \li{read_data}:
% \begin{enumerate}
% \item The language's syntax definition determines that \li{read_data(a)} is an expression of function application form.
% \item Similarly, \li{read_data} and \li{a} are definitively expressions of variable form.
% \item The variable \li{a} can only refer to the binding of \li{a} on Line 1.
% \item The variable \li{w} can be renamed without knowing anything about the values that \li{read_data} and \li{a} stand for.
% \item The type of \li{x} can be determined to be \li{B} by determining that the type of \li{read_data} is \li{A -> B} for some \li{A} and \li{B}, and checking that \li{a} has type \li{A}. Nothing else needs to be known about the values that \li{read_data} and \li{a} stand for. In Reynolds' words \cite{B304}:
% \begin{quote}
% \emph{Type structure is a syntactic discipline for enforcing levels of abstraction.}
% \end{quote}
% \end{enumerate}

\subsection{Term-Rewriting Systems Offer Limited Syntactic Control}
An alternative approach that a library provider might consider when seeking to control syntactic cost is to leave the context-free syntax of the language unmodified and instead contextually repurpose select existing syntactic forms using a \emph{term rewriting system}.

Na\"ive term rewriting systems suffer from problems analagous to those that plague syntax definition systems. In particular, there can be conflicts because separately defined rewriting rules might attempt to rewrite the same term differently. It can also be difficult to determine which rewriting rule, if any, is responsible for any particular term. Certain conventions have emerged, e.g. OCaml's textual syntax now includes \emph{preprocessor extension (ppx) points} used to identify terms that some presumably corresponding term rewriting preprocessor must rewrite \cite{ocaml-manual}, but a correspondence is not enforced. In any case, it is difficult to reason abstractly about types and binding given a program subject to a large number of context-free rewriting rules.

Modern \emph{term-rewriting macro systems}, however, have made some progress. In particular:
\begin{enumerate}[nolistsep]
\item Macro systems require that the client explicitly apply the intended rewriting (implemented by a macro) to the term that is to be rewritten, thereby addressing the problems of conflict and determining responsibility. However, it is often unclear whether a given macro transforms or reinterprets the form of a given argument or sub-term thereof, as opposed to treating it parametrically by inserting it unmodified into the generated expansion. This is closely related to the problem of determining a {segmentation} discussed above.
\item \emph{Hygienic} macro systems prevent variable {capture} and enforce {context-independence} \cite{Kohlbecker86a,DBLP:conf/popl/Adams15,DBLP:conf/popl/ClingerR91,DBLP:journals/lisp/DybvigHB92}.
\item Although much of the research on macro systems has been for languages in the Lisp tradition \cite{mccarthy1978history}, some progress has also been made on reasoning about types with the design of \emph{typed macro systems} where annotations constrain the macro arguments and the generated expansions. Examples include Scala's macro system \cite{ScalaMacros2013}, formal calculi studied by \citet{DBLP:conf/esop/HermanW08} and \emph{staging macro systems} like MetaML \cite{Sheard:1999:UMS} and MacroML \cite{ganz2001macros} (which, it is worth noting, do not give the macro access to the syntax tree of arguments.)
\end{enumerate}

Perhaps the biggest problem, then, is that term-rewriting systems offer library providers only limited syntactic control. In other words many syntactic forms of interest will not particularly resemble those available in the host language. For example, consider the  XHTML and K examples above. In both cases, the syntactic conventions are quite distinct from those of ML-like languages (and, for that matter, languages that use S-expression.)

It is tempting in these situations to consider repurposing string literal forms. For example, we might wish to apply a macro \li{html!} (following Rust's convention of using a post-fix \li{!} to distinguish macro names from variables) to rewrite string literals containing Ur/Web-style XHTML syntax as follows:
\begin{lstlisting}[numbers=none]
  fun heading first_name = html! 'SSTR<h1>Hello, {[first_name]}!</p>ESTR'
  val body = html! 'SSTR<body>{heading "World"} ...</body>ESTR'
\end{lstlisting}

The problem here is that there is no way to extract the spliced expressions from the supplied string literal forms while satisfying the hygiene conditions, because variables that come from these spliced terms (here, \li{first_name} and \li{heading}) are indistinguishable from variables that inappropriately appear free in the generated expansion. In addition, the problem of segmentation becomes even more pernicious: to a human or tool unaware of Ur/Web's syntax, it is not immediately apparent which particular subsequences of the string literals supplied to \li{html!} are segmented out as spliced expressions. Reader macros have essentially the same problem  \cite{DBLP:journals/jfp/FlattCDF12}.

  Some languages build in \emph{string splicing} (a.k.a. \emph{string interpolation}) forms, e.g. Scala \cite{odersky2008programming}, or similar but more general \emph{fragmentary quotation forms} \cite{conf/icfp/Slind91}, e.g. the SML/NJ dialect of ML. These designate particular delimiters to allow escape out into the expression language. The problem with using these (dynamically, or in conjunction with term rewriting macros) as vehicles to introduce syntactic sugar at various types is 1) there is no ``one-size-fits-all'' escape delimiter, and 2) reasoning about typing is problematic because every escaped term is checked against the same type. In the example above, we spliced in two different types of expressions using two different delimiters. In general, e.g. when defining syntax for a programming language with many sorts of terms, the most appropriate choice of delimiters might depend on where each spliced term appears.

% These existing forms normally have other meanings, so this can be confusing \cite{pane1996usability}.

\subsection{Contributions}
This work introduces a system of \textbf{typed syntax macros (TSMs)} that gives library providers substantially more syntactic control than existing typed term-rewriting macro systems while maintaining the ability to reason abstractly about types, binding and segmentation.% abstract reasoning principles. % comparable to the level of control they have when defining a syntax dialect.

% Clients apply TSMs to \emph{generalized literal forms}. 
Our take on the running example is shown in Fig. \ref{fig:first-tsm-example}: we apply a TSM named 
\li{$html} 
 to \emph{generalized literal forms} delimited by backticks. TSM names are prefixed by \li{$} 
 to clearly them from variables. The semantics delegates control over the parsing and expansion of each literal body to the applied TSM during a semantic phase called \emph{typed expansion}, which generalizes the usual typing phase. 
\begin{figure}[ht!]
\vspace{-10px}
\begin{lstlisting}
  fun heading first_name = $html `SURL<h1>Hello, {[EURLfirst_nameSURL]}!</h1>EURL`
  val body = $html `SURL<body>{EURLheading "World!"SURL} ...</body>EURL`
\end{lstlisting}
\vspace{-10px}
\caption{Two examples of a TSM being applied to a generalized literal form. %In each case, literal body, between backticks, is initially left unparsed according to the language's context-free syntax. %The applied TSM determines a segmentation and expansion during the typed expansion phase, which generalizes the usual typing phase. 
We assume strings are primitive.}
\vspace{-10px}
\label{fig:first-tsm-example}
\end{figure}

Generalized literal forms, which first arose in related work by \citet{TSLs} that we will detail in Sec. \ref{sec:related-work}, subsume a variety of common syntactic forms because the context-free syntax of the language only defines which outer delimiters are available (in this paper, we will mostly use backticks, but in general, there can be many choices.) \emph{Literal bodies}, i.e. the characters between the delimiters, are otherwise syntactically unconstrained. Choices regarding internal delimitation of spliced terms is not imposed by the language definition. Notice, however, that Figure \ref{fig:first-tsm-example} reveals the locations of the spliced expressions by coloring them black. We have designed our system so that a figure like this, which presents a \emph{segmentation} of each literal body into spliced terms (in black) and characters parsed in some other way by the applied TSM (in green)\footnote{In general, spliced expressions might themselves apply TSMs, in which case the convention is to use a distinct color for unspliced segments at each depth. For example, if strings were not primitive, we might write \li{`SURL<body>\{EURLheading ($str "SSTRWorld!ESTR")SURL\} ...</body>EURL`}.}, can always be  generated.

% Because the context-free syntax is never extended, syntactic conflicts are not a concern.

%As such, the semantics can take the type and binding structure of the surrounding program into account when validating the expansion that the TSM programmatically generates to ensure that clients can answer critical questions related to types and binding, like those enumerated in Section \ref{sec:abs-reasoning-intro}. Clients need not have knowledge of the implementation of the TSM or of the generated expansion, i.e. there are useful principles of syntactic abstraction.

% The primary technical challenge has to do with the fact that the applied TSM needs to be able to splice terms out of the literal body for inclusion in the expansion. For example,

 % We design our mechanism such that these locations can easily be determined from the output of the TSM. This is essential for our hygiene mechanism, and it is also useful in that this information can be presented to the user (e.g. as shown in Figure \ref{fig:first-tsm-example-marked}). %As such, we must develop a mechanism where 1) the positions of spliced subterms can be determined without examining the macro implementation (e.g. so that they can be presented to the user differently by an editor or pretty-printer, ;  and 2) the hygiene mechanism must give only portions of the expansion that correspond to these spliced subterms access to the application site context.

%In order to reason about types and binding, client programmers need only have knowledge of 1) the segmentation (e.g. by examining a figure like this presented by a code editor or pretty-printer) and 2) a type annotation on the definition of the applied TSM. No other details about the applied TSM's implementation or the expansion that it generates need to be revealed to the client programmer. In other words, 
TSMs come equipped with useful principles of syntactic abstraction. We will more precisely characterize these abstract reasoning principles as we proceed. For now, to develop some intuitions, consider Figure \ref{fig:K-tsm-example}, which uses TSMs to express the ``unreasonable'' example from Figure \ref{fig:K-dialect}. Without examining the expansion of Line 3, we can reason as follows:

\begin{enumerate}
\item \textbf{(Responsibility)} The applied TSM, \li{$kquery}, is solely responsible for typed expansion of the literal body. 
\item \textbf{(Segmentation)} By examining the segmentation, we know that the two instances of \li{x} on Line 3 are parsed as spliced expressions, whereas \li{R} is parsed in some other way peculiar to this form.
\item \textbf{(Capture)} The system prevents capture, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2 -- it cannot capture an unseen binding introduced in the expansion of Line 3.
\item \textbf{(Context Dependence)} The system enforces context independence, so the expansion of Line 3 cannot  rely on the fact that, for example, \li{w} is in scope.
\item \textbf{(Typing)} An explicit type annotation on the definition of \li{$kquery} determines the type that every expansion it generates will have. We will see an example of a TSM definition in the next section.

Moreover, each segment in the segmentation also comes paired with the type it is expected to have. This information is usually not necessary to reason about typing, but it can be conveyed to the programmer upon request by the program editor if desired. %TSM definitions follow the usual scoping rules, so it is easy to ``jump to the definition'' of \li{$kquery}.
\end{enumerate}

\begin{figure}
\begin{lstlisting}[numbers=none,xleftmargin=0px]
  val w = compute_w ()
  val x = compute_x ()
  val y = $kquery `SURL(!R)@&{&/EURLxSURL!/:2_!EURLxSURL}'!R}EURL`
\end{lstlisting}
\vspace{-5px}
\caption{TSMs make examples like the one from Fig. \ref{fig:K-dialect} more reasonable.}
\vspace{-5px}
\label{fig:K-tsm-example}
\end{figure}

% \begin{figure}[h]
% \begin{lstlisting}
% PElement Nil Seq(
%   TextNode "Hello, ", 
%   Seq(TextNode (join(" ", Cons(first, Cons(second, Nil)))), 
%   TextNode "!"))
% \end{lstlisting}
% \caption{The desugaring.}
% \end{figure}

% There is also no ambiguity with regard to which TSM has control over each form, and searching for the definition of a TSM is no more difficult than searching for any other binding, i.e. there are well-defined scoping rules.

% In other words, TSMs maintain a useful notion of syntactic abstraction. %More specifically, TSMs maintain a \emph{hygienic binding discipline}, meaning that questions Questions 4 and 5 above were concerned with are disallowed entirely. 
% We will, of course, make this notion more technically precise as we continue.

The remainder of this document is organized as follows.
\begin{itemize}
\item Sec. \ref{sec:setsms} begins by going through the mechanics of defining a {simple expression TSM} and describes the process of typed expansion and expansion validation by example.
\item Sec. \ref{sec:sptsms} does the same for a {simple pattern TSM}, i.e. a TSM that generates patterns that match values of a specified type. The hygiene condition for pattern TSMs is interesting.
\item Sec. \ref{sec:setsms-formally} defines a calculus of simple expression and pattern TSMs and formally establishes the reasoning principles implied above.

% In the remaining chapters, we enrich the language developed in Chapter \ref{chap:uetsms} with advanced features based on those found in full-scale general-purpose languages like ML and Scala, and enhance our TSM mechanism with these new features.

% We next develop a more sophisticated account of TSMs in Chapters \ref{chap:ptsms} and \ref{chap:static-eval}, with the aim of working out details necessary to integrate TSMs into full-scale general-purpose languages like ML or Scala. %These two chapters constitute Part \ref{part:parametric-tsms} of our contributions.
\item Sec. \ref{sec:ptsms} adds parameterized types and an ML-style module system and introduces \emph{parametric TSMs} (pTSMs), i.e. TSMs that take type and module parameters. Parameters serve two purposes:
\begin{enumerate}
\item They allow for TSMs that operate not just at a single type, but over a type- and module-parameterized family of types. For example, rather than defining a TSM % \li{$strlist} 
for string-valued dictionaries and another TSM 
%\li{$intlist} 
for integer-valued dictionaries, or over different implementations of the dictionary signature, we can define a single parametric TSM 
%\li{$list} 
that operates uniformly across a type- and module-parameterized family of dictionary types.
\item Parameters also give expansions access to terms other than those explicitly spliced in to the literal body. This is particularly convenient for working with abstract types \emph{a la} ML. Partial parameter application in TSM abbreviations lowers the cost of this explicit parameter-passing discipline.
\end{enumerate}
This section culminates with a calculus of ML-style modules that has support for parametric TSMs.
\item Sec. \ref{sec:static-eval} considers the topic of term evaluation during the typed expansion phase, i.e. \emph{static evaluation}, in a bit more depth, and also gives examples of TSMs that are useful for defining other TSMs, e.g. TSMs that implement parser generators and quasiquotation.
\item Sec. \ref{sec:related-work} discusses additional related work, including closely related work by \citet{TSLs,sac15} on type-specific languages. We conclude with a brief discussion of present limitations and future directions.
\end{itemize}

Due to space limitations, many of the standard or straightforward technical details of the calculi defined in Sec. \ref{sec:setsms-formally} through Sec. \ref{sec:static-eval} are relegated to the supplemental material. For convenience, the examples that we give are written in a full-scale functional language under development called VerseML. % For the present purposes, we can consider VerseML as essentially the calculus defined in the supplement extended with various conveniences that are commonly found in other ML-like languages and, notionally, orthogonal to TSMs. %VerseML is, as its name suggests, a conceptual descendent of ML. It diverges from other dialects of ML that have a similar type structure in that it has a bidirectional type system \cite{Pierce:2000:LTI:345099.345100} (like, for example, Scala \cite{OdeZenZen01}) for reasons that have to do with the mechanism of TSM implicits described in Chapters \ref{chap:tsls} and \ref{chap:ptsms}. 
%The reason we will not follow Standard ML \cite{mthm97-for-dart} in giving a complete formal definition of VerseML in this work is both to emphasize that the primitives we introduce are ``insensitive'' to the details of the underlying type structure of the language (so TSMs can be considered for inclusion in a variety of languages, not only dialects of ML), and to avoid distracting the reader (and the author) with definitions that are already well-understood in the literature and that are orthogonal to those that are the focus of this work. 
% We will not formally define these features mainly to avoid unnecessarily complicating our presentation with details that are not essential to the ideas presented herein. As such, 
All examples written in VerseML should be understood to be informal motivating material for the subsequent formal material. %We anticipate that future full-scale language specifications will be able to combine the ideas  in the proposed work without trouble. %The purpose of the work being proposed is to serve as a reference for those interested in the new constructs we introduce, not to serve as a language specification. 
%We will give a brief overview of these languages are organized in Sec. \ref{sec:VerseML}.

% \begin{figure}[h]
% \begin{lstlisting}[numbers=none,xleftmargin=0px]
% let syntax $strlist = $list string in 
% $html `SURL<p>Hello, {[EURLjoin ($str ' ') ($strlist [firstSURL,EURL last])SURL]}</p>EURL`
% \end{lstlisting}
% \caption{The example from Figure \ref{fig:first-tsm-example-marked}, expressed using a parametric TSM.}
% \label{fig:first-ptsm-example-marked}
% \end{figure}

% In Secs. \ref{chap:uetsms} and \ref{chap:uptsms}, we assume for the sake of technical simplicity that each TSM definition is self-contained, needing no access to libraries or to other TSMs. This is an impractical assumption in practice. We relax this assumption in In

% %\item \textbf{Type-specific languages}, or \textbf{TSLs}. TSLs, described 
% In Chapter \ref{chap:tsls}, we develop a mechanism of \emph{TSM implicits} that allows library clients to contextually designate, for any type, a privileged TSM at that type. The semantics applies this privileged TSM implicitly to unadorned literal forms that appear where a term of the associated type is expected. For example, if we designate 
% %\li{$str} 
% as the privileged TSM at the \li{string} type and 
% %\li{$strlist}
% as the privileged TSM at the \li{list(string)} type, we can express the example from Figure \ref{fig:first-tsm-example-marked} instead as shown in Figure \ref{fig:first-tsm-example-implicit} (assuming \li{join} has type \li{string -> list(string) -> string}.) 
% \begin{figure}[h]
% \begin{lstlisting}[numbers=none]
% $html`SURL<p>Hello, {[EURLjoin ' ' [firstSURL,EURL last]SURL]}</p>EURL`
% \end{lstlisting}
% \caption{The example from Figure \ref{fig:first-tsm-example-marked} drawn to take advantage of TSM implicits.}
% \label{fig:first-tsm-example-implicit}
% \end{figure}
